---
title: "Merging/Joining & Reshaping Data"
author: "Joshua F. Wiley"
date: "`r Sys.Date()`"
output: 
  tufte::tufte_html: 
    toc: true
    number_sections: true
---

```{r, echo=FALSE, include=FALSE}
## note this chunk is NOT required
library(checkpoint)
checkpoint("2019-01-25", R.version = "3.5.1",
           scanForPackages = FALSE)
```

Download the raw `R` markdown code here[^notenote]
[https://jwiley.github.io/MonashHonoursStatistics/Data_Pt2.rmd](https://jwiley.github.io/MonashHonoursStatistics/Data_Pt2.rmd).

[^notenote]: I am trying to separate out core ideas from additional
    notes/details. Core ideas and code are in the main document. Extra
    information that may help understanding but is not the big picture
    will go in margin notes, like this.

```{r setup}
options(digits = 2)

library(data.table)
library(reshape2)
library(ggplot2)
library(lme4)

## read in the dataset
d <- readRDS("aces_daily_sim_processed.RDS")

```

# SETUs

Please complete the SETUs for PSY4210 here:
[https://monash.bluera.com/monash](https://monash.bluera.com/monash).
Your feedback and evaluation is critical to understanding what worked
and where we need to improve in this unit. In honours, the class size
is much smaller. Every evaluation counts. Please help me by completing
it now.

# Merging Data

Often data come from multiple sources and are in separate files. For
example, a baseline and post intervention questionnaire may be
collected separately, or you may have questionnaire data and sleep
data from a FitBit. 

Data can be combined by merging, and merging in `R` is typically done
using the `merge()` function.

There are four main types of merges in `R` using the `merge()`
function[^joins]. Assume we are merging two datasets: `x` and
`y`[^xy]. The type of merge performed is controlled by arguments to
the `merge()` function.

[^joins]: Merging in `R` is the same as "joins" which is the more
common term for databases, such as SQL.

[^xy]: Merging or joins always involves two datasets. Even if you need
to merge more than two datasets (e.g., you have baseline, post, and
follow-up data) you always merge / join them two at a time. For
example, first joining baseline and post into one `bp` dataset and
then joining `bp` and your follow up data. In `R` the `x` dataset is
always the "left" and the `y` dataset is always the "right".
	
- **Natural join**: resulting data has only rows present in
  **both** `x` and `y`. Argument: `all = FALSE` .
- **Full outer join**: the data has all rows in `x` and all rows
  in `y`. Argument: `all = TRUE` .
- **Left outer join**: resulting data has all rows in `x`. Argument: `all.x
  = TRUE` .
- **Right outer join**: resulting data has all rows in `y`. Argument: `all.y
  = TRUE` .

We will explore joins using two, small datasets[^data]. `ID` is used
to indicate participant ID.

[^data]: This small, hypothetical study collected data via surveys and
actigraphy. The surveys have people's ages and the actigraphy has
their sleep duration, which we wish to join together for analysis.

```{r}

surveys <- data.table(
  ID = c(1, 2, 3),
  Age = c(19, 18, 20))

acti <- data.table(
  ID = c(2, 3, 4),
  Sleep = c(8, 6, 7))

## view the two small datasets
print(surveys)
print(acti)

```

## Natural Join in `R`

**Only rows present in both datasets**[^naturaljoin]. 

Joining data requires specifying how `R` should know which rows match
each other. This is typically  done by an ID variable using the code:
`by = "ID"`[^byid]. The argument, `all = FALSE` is what tells `R` this should
be a natural join[^naturaljoin2].

[^naturaljoin]: Natural joins have only the rows / observations that
are present in both datasets. In this example, the surveys and the
actigraphy. You can pick whichever dataset you want to be `x` and 
and `y` by which you list first.

[^byid]: When we write `by = "ID"` what is meant is that it should be
    by the variable called "ID". The word ID is not special, its just
    that we happened to call the variable containing our IDs, "ID". If
    we had called the variable containings IDs "Email" or "Name" then
    we would write: `by = "Email"` or `by = "Name"`. In other words,
    match the text in quotes to an actual variable name in the data.

[^naturaljoin2]: I.e., a join with only the rows present in both `x`
and `y` datasets actually returned.

```{r}

merge(
  x = surveys,
  y = acti,
  by = "ID",
  all = FALSE)

```

## Full Outer Join

**All rows present in either dataset**[^fullouter]. 

The argument, `all = TRUE` is what tells `R` this should be a full
outer join[^joinmissing].

[^fullouter]: Full outer joins include all rows or  observations that
are present in either dataset. In this case, in either the surveys
or the actigraphy. You can pick whichever dataset you want to be "x"
and "y" by which you list first, although with a full outer join, it
does not make much difference, other than which columns/variables come
first and which come second. As before, we must specify what variable(s) to
merge by so `R` knows how to match rows together, accomplished using
the code: `by = "ID"`. 

[^joinmissing]: Note that `R` fills in missing values for rows /
observations and variables that are present in one dataset but not the
other. So if two datasets do not have identical rows / observations, a
full outer join will result in some missing values. These are shown in
the print out as `NA`.

```{r}

merge(
  x = surveys,
  y = acti,
  by = "ID",
  all = TRUE)

```


## Left Outer Join

**All rows / observations that are present in left dataset
(surveys)**[^leftouter].

The argument, `all.x = TRUE` is what tells `R` this should be a left
outer join.

[^leftouter]: As before, we must specify what variable(s) to
merge by so `R` knows how to match rows together, accomplished using
the code: `by = "ID"`. Also as before, `R` fills in missing values for
the variables. In this case, missing values for the rows /
observations that are present in surveys, but not in acti (i.e., the
variable Sleep and ID = 1).

```{r}

merge(
  x = surveys,
  y = acti,
  by = "ID",
  all.x = TRUE)

```

## Right Outer Join

**All rows / observations that are present in right dataset
(acti)**[^rightouter]. 

The argument, `all.y = TRUE` is what tells `R` this should be a right
outer join.

[^rightouter]: As before, we must specify what variable(s) to merge by
so `R` knows how to match rows together, accomplished using the code:
`by = "ID"`. Note that `R` fills in missing values for the variables
in surveys that are not in acti and the rows / observations that are
present in acti, but not in surveys 
(i.e., the variable Age and ID = 4).

```{r}

merge(
  x = surveys,
  y = acti,
  by = "ID",
  all.y = TRUE)

```

## Rows and missing after joining

It is helpful to know how many rows you expect in the results.

Following are two new small datasets. In each, ID 2 is repeated
twice. One might anticipate that a full outer join would return 5
rows, (IDs 1, 2, 2, 3, 4). However, `R` actually returns 7 rows, with
four rows for ID 2, because all rows of ID 2 in `surveys2` are repeated by
all rows of ID 2 in `acti2`. **Merging with any duplicates can behave
in unexpected ways**. In a small dataset, it is easy to see what has
happened. In a large dataset, checking whether the merge worked as
expected requires using summaries or other checks, such as evaluating
whether the number of rows matched what you wanted or needed.

```{r}

surveys2 <- data.table(
  ID = c(1, 2, 2, 3),
  Age = c(19, 18, 18, 20))

acti2 <- data.table(
  ID = c(2, 2, 3, 4),
  Sleep = c(8, 7, 6, 7))

## view small datasets
print(surveys2)
print(acti2)

merge(surveys2, acti2, by = "ID", all = TRUE)

```


## You Try It - Merging

```{r datamerge}

ds1 <- data.table(
  ID = c("A1", "B2", "C3"),
  Stress = c(4, 5, 6))

ds2 <- data.table(
  ID = c("A1", "B2", "D4", "E5"),
  Steps = c(9524, 15303, 7432, 4035))

``` 

Here are those two datsets shown as tables. `ds1`

```{r, echo = FALSE, results = "asis"}
knitr::kable(ds1, caption = "Dataset: ds1")
```

and `ds2`

```{r, echo = FALSE, results = "asis"}
knitr::kable(ds2, caption = "Dataset: ds2")
```

Before you try merging those datasets, see if you could answer these
questions:

- Assuming `ds1` is the "left" dataset and `ds2` is the "right"
  dataset, how many rows would occur in a **natural join** of `ds1` and
  `ds2`? How many missing values total across all variables?

- Assuming `ds1` is the "left" dataset and `ds2` is the "right"
  dataset, how many rows would occur in a **full outer join** of `ds1` and
  `ds2`? How many missing values total across all variables?

- Assuming `ds1` is the "left" dataset and `ds2` is the "right"
  dataset, how many rows would occur in a **left outer join** of `ds1` and
  `ds2`? How many missing values total across all variables?

- Assuming `ds1` is the "left" dataset and `ds2` is the "right"
  dataset, how many rows would occur in a **right outer join** of `ds1` and
  `ds2`? How many missing values total across all variables?

These are the sort of questions that would be on the exam.

```{r trymerge, error = TRUE}

## using the datasets ds1 and ds2, do a natural join
merge(   )


## using the datasets ds1 and ds2, do a full outer join
merge(   )


## using the datasets ds1 and ds2, do a left outer join
merge(   )


## using the datasets ds1 and ds2, do a right outer join
merge(   )


``` 

## Merging Data by Multiple IDs

Sometimes a single variable is not enough to merge two datasets. This
is particularly true in longitudinal studies where there are repeated
measures. `R` can handle this by using multiple variables as IDs for
merging. Alternately, a second, unique ID can be created by combining
two or more pieces of information. The following shows an example.

```{r}

surveys <- data.table(
  ID = c(1, 1, 2, 2),
  Time = c(1, 2, 1, 2),
  Stress = c(6, 4, 3, 4))

acti <- data.table(
  ID = c(1, 1, 1, 3, 3),
  Time = c(1, 2, 3, 1, 2),
  Sleep = c(7, 8, 9, 8, 8))

## incorrect
merge(surveys, acti,
      by = "ID",
      all = TRUE)

## correct
merge(surveys, acti,
      by = c("ID", "Time"),
      all = TRUE)

``` 

## Merging Data by Different IDs

Often data are under our control, but sometimes, they are not or even
if they are ID variables are not named consistently across datasets.
As long as the actual ID values match, the variable names do not need
to be the same. For example, in Qualtrics, if you create custom links,
IDs may be called "RecipientFirstName" while on the actigraphy data,
you may have called IDs, "ID". `R` can handle this easily by using the
arguments, `by.x` and `by.y` to give the variable names in the `x` and
`y` datasets, respectively.


```{r}

surveys2 <- data.table(
  FirstName = c(1, 2, 2, 3),
  Age = c(19, 18, 18, 20))

acti2 <- data.table(
  ID = c(2, 2, 3, 4),
  Sleep = c(8, 7, 6, 5))

merge(
  x = surveys2,
  y = acti2,
  by.x = "FirstName",
  by.y = "ID",
  all = TRUE)

```

## Checking and Preparing Data for Merging

Because merging data can have unintended consequences if matched
incorrectly, often it is a good practice to check the data in advance
to make sure it is what you expect.

Generally, one-to-one merges, where each row of one dataset is matched
to one and only one row of another dataset is easiest. You can have
duplicates and these merge well in two cases: (1) the left, `x`
dataset has repeated measures but the right, `y` dataset only has one
row of data per ID or (2) the left, `x` dataset only has one row of
data per ID but the right, `y`, dataset has repeated measures per ID.
Many to many merges, involves *repeated IDs* in both datasets being
merged often is not what is intended. Note that repeated IDs does not
mean repeated measures. You can merge two repeated measures datasets
but not have repeated IDs by using two variables as your ID variable
or by creating unique IDs that combine ID and time, for example.

To check data, there are a few helpful functions in `R`.
The `anyDuplicated()` function returns 0 if no values are duplicated
and the position where the first duplicate occurs if any values are
duplicated.

```{r}

## there are duplicate IDs, starting in position 2
anyDuplicated(acti2$ID)

## there are no duplicated sleep values
anyDuplicated(acti2$Sleep)

## there are duplicated IDs here too
anyDuplicated(acti$ID)

## however if we paste / concatenate
## both ID and time, there are no duplicates
anyDuplicated(paste0(acti$ID, acti$Time))

```

If you do not expect duplciates, it is a good idea to check for them
before merging, as `R` will not tell you if there are duplicates and
it impacted the merge.

Relatedly, if there are duplicates, you can count how many there are,
using the `duplicated()` function and making a frequency table using
`table()`. 

```{r}

table(duplicated(acti2$ID))

``` 

Another way to approach this is to create a frequency table of IDs and
then create a frequency table of the frequencies. If everyone only
shows up once, you should have many frequencies of 1. If any ID shows
up more than once, you will have non zero frequencies > 1.
The frequency table below, shows that two IDs showed up once and one
ID showed up two times.

```{r}

table(table(acti2$ID))

``` 

Other checks are the number of unique IDs in each dataset. This is
accomplished by using the `unique()` function to first exclude any
duplicated values and then finding the length of the IDs.

```{r}

length(unique(surveys2$FirstName))

length(unique(acti2$ID))

```

You can also check how many IDs from one dataset are in another
dataset by combining the `unique()` function with the `%in%` logical
operator to get a series of TRUE or FALSE values if an ID from one
dataset is in another dataset. In large datasets, this may be many
results, so instead, we can create a frequency table of the results
using `table()`. Note that which set of IDs is on the left or the
right can make a difference. If you expect one dataset to contain all
IDs (e.g., you are merging a new variable into a large "master"
dataset that you believe has every single one of your participants)
this is a very helpful way to check whether that is true.

```{r}

table(unique(surveys2$FirstName) %in% unique(acti$ID))

table(unique(acti$ID) %in% unique(surveys2$FirstName))

``` 

## Checking Data You Try It

Using our daily data:

- Count how many unique IDs there are
- Create a frequency table of the number of times each `UserID` is
  duplicated.
- Check if there are duplicates if you use both `UserID` and
  `SurveyDay` to create a combined ID.
- Check if there are duplicates if you use `UserID` and `SurveyDay`
  and `Survey`.

```{r trychecking}

## you try to write the checking code here!



``` 

## Merging You Try It 2

Let's say that one participant had technical problems with the surveys
from the daily data we have worked with and so filled out the last few
on paper and pencil. Now we want to merge this data into our master
daily dataset. The resulting dataset, `d2` should have 6,930 rows.

```{r trymerging2, error=TRUE}

pencil <- data.table(
  UserID = 1L,
  SurveyDay = as.Date("2017-3-08"),
  Survey = c("Morning", "Afternoon", "Evening"),
  STRESS = c(1, 3, 1))

d2 <- merge(   )

``` 

# Reshaping Data

Another common data management task is reshaping data. This typically
occurs when data are stored in one format, such as different columns
for different time points, and you need data reshaped into another
format, such as multiple rows for each ID for mixed models.
What follows is a small sample of a "wide" dataset. That is, a dataset
where each timepoint is stored in a separate variable.

```{r}

dwide <- data.table(
  ID = c(1, 2, 3),
  STRESS1 = c(2, 1, 3),
  STRESS2 = c(4, 5, 6),
  STRESS3 = c(1, 1, 2))

``` 

```{r, echo = FALSE, results = "asis"}
knitr::kable(dwide, caption = "Example 'wide' dataset where different time points are stored in separate variables.")
```

`R` has a function, `reshape()` that allows reshaping data from wide
to long. It is quite complicated, so may not all be clear at first,
but its easiest to learn through examples. For now, we will focus on
five arguments to `reshape()`.

- `data` this is the dataset to reshape
- `varying` this is a list where each element is a vector of variable
  names that vary over time that belong to the same overall measure.
- `timevar` this is the name of the variable indicating to which time
  a measure belong, it will be a new variable in a wide to long
  reshape.
- `idvar` this is the name of the ID variable in the data and is
  required.
- `direction` this is the direction to reshape ("long" or "wide").

```{r}

reshape(
  data = dwide,
  varying = list(
    STRESS = c("STRESS1", "STRESS2", "STRESS3")
  ),
  timevar = "time",
  idvar = "ID",
  direction = "long")

```


notes better on instructions, notes on how long to wait
