---
title: "Working with Different Data Types"
author: "Joshua F. Wiley"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_book: default
---

```{r, echo=FALSE, include=FALSE}
library(checkpoint)
checkpoint("2019-01-25", R.version = "3.5.1",
           scanForPackages = FALSE)
```

```{r setup}
options(digits = 2)

library(data.table)
library(chron)
library(ggplot2)
library(lme4)

## read in the dataset
d <- readRDS("aces_daily_sim_processed.RDS")

```

# Data Types

There are many types of data. In `R`, you can find what type of data a particular variable is by asking for its `class()`.

```{r}

class(d$Age)
class(d$BornAUS)
class(names(d))

``` 

- *Logical*: Used for logical data, which are either TRUE or FALSE. If data are logical, it is a very efficient format and useful for many cases. Logical variables can be compared. Arithmetic can be used for logical variables, in which case TRUE is treated as 1 and FALSE as 0.
- *Integer*: Used for integer type data, that is whole numbers like 0, 1, 2. For variables that are only whole numbers, integer format is more efficient than real numbers or numeric data (e.g., 1.4).
- *Numbers/real/numeric*: Used for real numbers, such as 1.1, 4.8. It also can be used for integer data (i.e., whole numbers only) but is a less efficient format. In `R` these are represented by the class `numeric` abbreviated `num`.
- *Text/character/string*: Used for text type data, such as names, qualitative data, etc. Also, any numbers can be stored as strings. In `R` these are represented by the class `character` abbreviated `chr`. Character data do not work with arithmetic operators, but can be sorted (e.g., alphabetically).

Although these are the basic building blocks of data, there also are special formats that can be built off of these, often off of numeric type data.

- *Date*: A special representation of numeric data when the data are dates. In `R` represented by multiple classes, including `Date`, `POSIX` and `chron`. The standard approach is to count the days since some origin. In `R`, the default origin in 1970-01-01. So 1970-01-01 is 0. 2 would indicate 1970-01-03. You can do some arthmetic on dates. For example adding or subtracting numbers, calculating the differece between dates.
- *Date and time*: A special representation of numeric data when the data are dates and times. In `R` represented by multiple classes, including `POSIX` and `chron`. The standard approach is to count the seconds since some origin. In `R`, the default origin in 1970-01-01 00:00:00. So 1970-01-01 00:00:00 is 0. 2 would indicate 1970-01-01 00:00:02. You can do some arthmetic on datetime objects. For example adding or subtracting numbers, calculating the differece between dates and times.
- *time*: A special representation of numeric data when the data are times. In `R` represented by class `chron`. 24 hours are formated so that they go from 0 to 1. For example, 3:31pm would be converted to 24 hour time: 15:30 and then to decimals for hours: 15.5 then converted ito range from 0 to 1: 15.5/24.
- *factor*: A special representation of numeric data when the data are fundamentally discrete. For example, if study condition is coded as 0 = control, 1 = medication, 2 = psychotherapy. The data could be stored numerically, or as text. If stored as a factor, R will do some convenient things automatically, such as dummy coding in regression models.


# Date Variables

Often when dates are read into R, they are character data.
We will work with some different character formats and 
convert them to date class in `R`.

```{r, error = TRUE}

## sample character data
x1 <- c("2019-03-12", "2017-09-15")
x2 <- c("2019-Mar-12", "2017-Sep-15")
x3 <- c("12/3/2019", "15/9/2017")
x4 <- c("12/3/19", "15/9/17")

# these are character strings
class(x1)


## convert character strings to Date classes
as.Date(x1, format = "%Y-%m-%d")
as.Date(x2, format = "%Y-%b-%d")
as.Date(x3, format = "%d/%m/%Y")
as.Date(x4, format = "%d/%m/%y")

## save results
d1 <- as.Date(x1, format = "%Y-%m-%d")

## look at the class once converted
class(d1)

## arithmtic on dats
d1 + 5 ## addition
d1 - 7 ## subtract a week
d1 * 5 ## multiplication does not make sense for dates, an error

## extract just one date
d1[1]
d1[2]

## compare two dates
d1[1] - d1[2]

## compare two dates and convert to a number
as.numeric(d1[1] - d1[2])

## compare two dates and convert to a number
## and then to years
as.numeric(d1[1] - d1[2])/365.25


``` 

To convert date into a date class in `R`, we need to tell it the format the dates are written in. This is done by specifying a "template".

- "%Y" means 4 digit year [upper case]
- "%y" means 2 digit year [lower case]
- "%m" means digit month
- "%b" means abbreviated month (e.g., "Jan")
- "%d" means digit day

These pieces can be re-ordered an combined and use any separator. For example "%Y-%m-%d" for YYYY-mm-dd or "%Y/%m/%d" for YYYY/mm/dd or "%Y %m %d" for YYYY mm dd or "%d %m-%Y" for dd mm-YYYY. This flexibility is needed as different people and countries commonly write dates in many different formats.

**Now you try to convert the following character strings into date class objects in `R` by fillng in the correct template in quotes for the `format = ` argument. If you do it incorrectly, you will get errors or `NA` indicating the text could not be converted to a date uing the template you supplied.

```{r youtry1, echo=TRUE, eval=FALSE}

y1 <- c("5/28/2018", "6/30/2019")
as.Date(y1, format = "") ## you complete

y2 <- c("28 01 2018", "15 06 2019")
dy2 <- as.Date(y2, format = "") ## you complete
dy2 ## show results

y3 <- c("28-1-18", "15-6-19")
as.Date(y3, format = "") ## you complete

## how many days are there between 28 January 2018 and 15 June 2019?
## use the object dy2 to complete like we did above



```


Dates have a variety of helpful functions that let you operate on or learn moe about them. We can use the `weekdays()`, `months()`, and `quarters()` functions to find out the day of week, month or quarter a particular date falls in. They return character strings.

```{r}

weekdays(d1)
months(d1)
quarters(d1)

```

The `chron` package also makes another function available, `is.weekend()` which returns logical TRUE/FALSE values for whether a particular date is a weekend or not. This can be useful, for example to test if positive affect is higher on weekends then weekdays, in the daily study we worked with.

```{r}

## neither date is a weekend
is.weekend(d1)

## is positive affec higheron weekends?
summary(lmer(PosAff ~ is.weekend(SurveyDay) + 
               (1 | UserID), 
             data = d))

``` 


Dates also can be compared to determine if one is greater than or less than another. This can be useful for selecting observations that meet specific criteria. For example, suppose we wanted to exclude any dates that occurred before the semester started. Not everyone in the daily diary study is a student, but if they were, we might expect positive affect to be different on break than during semester. Suppose semester started 1 March 2017, we might want to only include observations that fell on or after that date.
Note here we do not need to specify a template because `R` will assume YYYY-mm-dd when written this way.

```{r}

## summary of dates
summary(d$SurveyDay)

## select surveys where the survey date is greater than or equal to
## 2017-03-01
d2 <- d[SurveyDay >= as.Date("2017-03-01")]

## summary of dates
summary(d2$SurveyDay)

``` 

Now you try to create a dataset that excludes any weekends

```{r youtry2, echo = TRUE, eval = FALSE}

## complete to make a dataset that 
## excludes any weekend days
d3 <- d[ ]

## fit intercept model
summary(lmer(PosAff ~ 1 + 
               (1 | UserID), 
             data = d3))

```

# Time Variables

Time variables also are commonly read into `R` as text.
The `chron` package has functions to help with time data.

```{r}

t1 <- c("9:15:08", "13:09:00")
ty1 <- chron(times. = t1, format = c(times = "h:m:s"))
ty1

## add one hour
ty1 + (1/24)

## add 20 minutes
ty1 + (20/60/24)

```

Sometimes, you get times that are incomplete. `R` represents time down to at least seconds, so seconds ae required. Even if not available, you must *choose* seconds.

```{r, error = TRUE}

t2 <- c("9:15", "13:09")
## this will give an error
chron(times. = t2, format = c(times = "h:m"))
chron(times. = t2, format = c(times = "h:m:s"))

## we need to add seconds
t2b <- paste0(t2, ":00")
t2b

## now convert to a time
y2b <-chron(times. = t2b, format = c(times = "h:m:s"))
y2b

## calculate the time difference
y2b[2] - y2b[1]

## numeric as a fraction of 24 hurs
as.numeric(y2b[2] - y2b[1])

``` 

# Date and Time Variables

The `chron` package and function can be used for date and time variables.

```{r}

dates1 <- c("2019-05-24", "2019-06-14")
times1 <- c("19:20:17", "9:15:33")

dt1 <- chron(dates. = dates1, times. = times1,
      format = c(dates = "y-m-d", times = "h:m:s"))
dt1

dt1 + 1 ## add one day
dt1 + (1/24) ## add one hour

```

# Factor Variables

Factor variables can be made from numeric or 
integer or character data, but are converted into 
the factor format to tell `R` that they are discrete
variables. This is done using the `factor()` function.

```{r}

## some sample data
x <- c(0, 1, 2, 1, 1, 2, 0)

## convert to a factor in R
## adding specific labels
y <- factor(x, 
  levels = c(0, 1, 2),
  labels =c("Control", "Medication", "Psychotherapy"))
y


```

A helpful feature of factors is that `R` dummy 
codes them automatically in regression models.

```{r}

## regression model with cyl as a numeric variable
summary(lm(mpg ~ cyl, data = mtcars))

## frequency table
table(mtcars$cyl)

## mke a factor version of cyl
mtcars$cylF <- factor(mtcars$cyl,
  levels = c(4, 6, 8))

## regression with dummy coding
summary(lm(mpg ~ cylF, data = mtcars))

```


Now try using factors for dummy coding. In the daily study,
there is a varible, SurveyInteger. 1 = morning, 2 = afternoon,
3 = evening. Make it a factor and then fit a LMM.

```{r youtry3, eval = FALSE, echo = TRUE}

d[, SurveyIntegerF := factor(
  SuveyInteger,
  levels = c(    ),
  labels = c(    ))]

summary(lmer(PosAff ~ SurveyIntegerF + 
               (1 | UserID),
             data = d))

```







