---
title: "Graphs Part 1"
author: "Joshua F. Wiley"
date: "`r Sys.Date()`"
output: 
  tufte::tufte_html: 
    toc: true
    number_sections: true
  tufte::tufte_handout: 
    toc: true
    number_sections: true
  word_document:
    toc: false
---

```{r, echo=FALSE, include=FALSE}
## note this chunk is NOT required
library(checkpoint)
checkpoint("2019-01-25", R.version = "3.5.1",
           scanForPackages = FALSE)
```

Download the raw `R` markdown code here
[https://jwiley.github.io/MonashHonoursStatistics/Graphs_Pt1.rmd](https://jwiley.github.io/MonashHonoursStatistics/Graphs_Pt1.rmd).


```{r setup}
options(digits = 2)

## two new packages are: (1) cowplot and (2) ggthemes
## You can try installing
## by uncommenting the install.packages() code below

#install.packages("cowplot", type = "binary")
#install.packages("ggthemes", type = "binary")


## some people also report type="binary" does not work
## if that happens to you, try:
# install.packages("cowplot")
# install.packages("ggthemes")

## once installed, open the packages using the 
## library() calls below

library(data.table)
library(ggplot2)
library(cowplot)
library(ggthemes)

## read in the dataset
d <- readRDS("aces_daily_sim_processed.RDS")

## small sample dataset
dt <- data.table(
  Age = c(20, 30, 40, 50, 60),
  Memory = c(7, 5, 6, 4.5, 4))

## convert a few variables in mtcars dataset
## into discrete factor variables
mtcars$cyl <- factor(mtcars$cyl)
mtcars$vs <- factor(mtcars$vs)
mtcars$am <- factor(mtcars$am)

```

# Grammar of Graphics

`ggplot2` is based on the **g**rammar of **g**raphics, a framework for
creating graphs.

The idea is that graphics or data visualization generally can be
broken down into basic low level pieces and then combined, like
language, into a final product.

Under this system, line plots and scatter plots are essentially the
same. Both have data mapped to the x and y axes.  The difference is
the plotting symbol (**ge**ometries labelled `geom`s in `R`) in
is a point or line. The data, axes, labels, titles, etc. may be
identical in both cases.

`ggplot2` also uses aesthetics, which control how geometries are
displayed. For example, the size, shape, colour, transparency level
all are **aes**thetics.

# Univariate Graphs

To begin with, we will make some simple graphs using `ggplot2`. The
first step is specifying the dataset and mapping variables to axes.
For basic, univariate plots such as a histograms, we only need to
specify the dataset and what variable is mapped to the x axis.
We can re-use this basic setup with different **geom**etries to make
different graphs.

```{r}

pb <- ggplot(
  data = mtcars,
  aes(x = mpg))

```

Using our basic mapping, we can "add" a histogram geometry to view a
histogram[^histogram].

[^histogram]: Histograms define equal width bins on the x axis and
count how many observations fall within each bin. Bars display these
where the width of the bar is the width of the bin and the height of
the bar is the count (frequency) of observations falling within that
range. Histograms show a univariate distribution.

```{r, fig.width = 6, fig.height = 5, fig.cap = "A histogram in ggplot2 for mpg"}

pb + geom_histogram()

```

We also can make a density plot, which also attempts to show the
distribution, but using a smooth density function rather than binning
the data and plotting the frequencies. Like histograms, the height
indicates the relative frequency of observations at a particular
value. Density plots are designed so that they sum to one[^density]. 

[^density]: Density plots attempt to provide an empirical
approximation of the probability density function (PDF) for data.
A probability density function always sums to one (i.e., if you
integrated to get the area under the curve, it would always be one).
The underlying assumption is that the observed data likely come form
some relatively smooth distribution, so typically a smoothing kernel
is used so that you see the approximate density of the data, rather
than seeing exactly where each data point falls.
Density plots show a univariate distribution.

```{r, fig.width = 6, fig.height = 5, fig.cap = "A density plot for mpg"}

pb + geom_density()

``` 

Another type of plot are (stacked) dotplots. These are very effective
at showing raw data for small datasets. Each dot represents one
person. If two dots would overlap, they are stacked on top of each
other. While these often are difficult to view with large datasets,
for small datasets, they provide greater precision than histograms
[^dotplot].

[^dotplot]: Dot plots show the raw data. The data values are on the x
axis. If two data points would overlap, they are vertically
displaced leading to another name: stacked dot plots. They are good
for small datasets. The y axis is kind of like a discrete density. Dot
plots show a univariate distribution.

```{r, fig.width = 6, fig.height = 3, fig.cap = "A dot plot for mpg"}

pb + geom_dotplot()

```

Finally, we can make Q-Q plots, although these require sample values
instead of values on just the x-axis. We use the `scale()` function to
z-score the data on the fly. Finally, we add a line with an intercept
(`a`) and slope (`b`) using `geom_abline()` which is the line all the
points would fall on if they were perfectly normally distributed.

```{r}

ggplot(mtcars, aes(sample = scale(mpg))) +
  geom_qq() +
  geom_abline(intercept = 0, slope = 1)

```


## Mapping Additional Variables

We can map additional variables to aesthetics such as the colour to
include more information.  To break the dataset down further, it is
helpful to have a bit larger dataset than the 32 cars in `mtcars`. The
`iris` dataset has measurements on 150 flowers from three species of
iris. For density plots, separating by colour is easy, by adding the
`Species` variable as an additional aesthetic[^multidensity].

[^multidensity]: When there are multiple univariate distributions to
view, density plots are probably one of the most efficient
ways. Histograms are difficult to view because they either are
stacked, which makes interpretation more difficult or dodged which is
visually difficult to see, or overplotted, which can hide some of the
data.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Density plot coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, colour = Species)) +
  geom_density()

```

For histograms, rather than control the colour of the lines, it is
more helpful to control the fill colour. By default, overlapping bars
are stacked on top of each other.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Histogram coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_histogram()

```

Overlapping bars also can be dodged instead of stacked.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Dodged histogram coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_histogram(position = "dodge")

```

Multiple geometric objects can be combined. The following figure
combines dot and density plots into one[^densitydot]. 

[^densitydot]: Note that when combining density and dot plots the
scales do not always work out well. For example, if the range of the
dotplot on the y axis is 0 to 1 and the range of the density plot y
axis is 0 to .005, the two will not be easily plotted on the same
figure. When they are more similar, however, they can be.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Density and dot plot plot coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, colour = Species)) +
  geom_dotplot() + geom_density()

```

While the colour works well for the density plots, filling would work
better for the dotplot. We can map separate aesthetics to each
geometric object.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Density and dot plot plot coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length)) +
  geom_dotplot(aes(fill = Species)) +
  geom_density(aes(colour = Species))

```

## Try It - Univariate

Using the `mtcars` dataset:

1. make a histogram for the variable: `qsec`.
2. make a QQ plot for qsec
3. Make a dotplot and density plot for: `disp` *separated* 
   by `am`[^tryitnote]:
   
[^tryitnote]: The dotplot and density plot do not show up well
together with these particular data as the density plot has very small
y axis values, relative to the dotplot.

```{r tryunivariate, error=TRUE}

## histogram code here (qsec)

## QQ plot code here (qsec)

## dotplot and density plot code here (disp by am)


```

# Bivariate Graphs

We can make bivariate plots by mapping variables to both the x and
y-axis. For a scatter plot, we use point geometric objects.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot of mpg and hp"}

ggplot(mtcars,
       aes(
         x = mpg,
         y = hp)) +
  geom_point()

```

We also can use lines for bivariate data. For this example, we will
use a small made up dataset. Compared to a scatter plot, we only
change point to line geometric objects.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Line plot of age and memory"}

ggplot(dt, aes(Age, Memory)) +
  geom_line()

```

With relatively discrete x axis, we can use a barplot for bivariate
data. By default, `geom_bar()` calculates the count of observations
that occur at each x value, so if we want our values to be the actual
bar height, we set `set = "identity"`.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity")

```

The grammar of graphics is designed to be like sentences, where you
can add or modify easily. For example, "There is a ball." or "There is
a big, red, striped ball." are both valid sentences. So to with
graphics, we often can chain pieces together to make it more nuanced.
In `R` we just "add" more by separating each component with `+`. 
Note that most argument names (i.e., `data = `, `mapping = `) are not
strictly required. `R` will match input to the correct argument by
position, often. The two sets of code below yield the same plot. In
the first, we explicitly label all arguments, in the second we rely on
position matching. Positional matching does not always work, for
example we still must specify `size = ` because we don't always
provide input for every argument, instead relying on defaults and only
changing the specific arguments we want changed from defaults.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory"}

ggplot(
  data = dt,
  mapping = aes(
    x = Age,
    y = Memory)) +
  geom_bar(
    mapping = aes(fill = Age),
    stat = "identity") +
  geom_line(size = 2) + 
  geom_point(size = 6)

ggplot(dt, aes(Age, Memory)) +
  geom_bar(aes(fill = Age), stat = "identity") +
  geom_line(size = 2) + 
  geom_point(size = 6)

```

# Improving Data Visualization

Before we continue examining graphs, it is helpful to think a bit more
about what makes a good graph or good data visualization.

Edward Tufte and William Cleveland are two authors who have written
extensively on data visualization and how to make good graphs. There
work is well worth reading to improving understanding on how to
efficiently convey data graphically.

- Tufte: [https://www.edwardtufte.com/tufte/](https://www.edwardtufte.com/tufte/)
- Cleveland: [http://www.stat.purdue.edu/~wsc/](http://www.stat.purdue.edu/~wsc/)

A few additional pages with good discussions on making good graphs:
[Graphic Design Stack Exchange 1](https://graphicdesign.stackexchange.com/questions/35052/how-to-visualise-two-dimensional-scientific-data-points-in-a-chart-in-graysca/35062#35062)
and [Graphic Design Stack Exchange 2](https://graphicdesign.stackexchange.com/questions/36908/best-plotting-symbols-for-scientific-plots-with-multiple-datasets/57122)

One key principle that Tufte emphasises is the data to ink ratio. This
ratio is how much data is conveyed versus ink used, and Tufte argues
to try to maximize this (i.e., more data, less ink). To see this,
consider the following graph, which is a fairly standard way stats
programs (Excel, etc.) tend to make barplots.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory - Step 1"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity") +
  theme_bw()

``` 

For starters, the borders tell us nothing. They edge the space but
convey no information. This can be cleaned up using a different
theme. Once we loaded the `cowplot` package, it automatically set its
own theme as the default, so if we leave off the `theme_bw()` we get a
cleaner graph.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory - Step 2"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity")

``` 

However, there are still some borders, which we can strip away with no
loss of data, but reducing the ink.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory - Step 2"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity") +
  theme(axis.line = element_blank())

``` 

Next, think about what data are conveyed in this graph. The bars
capture two pieces of information: (1) the age and (2) the memory at
that age. The only pieces of the bars we really need are the top. The
rest of the bars take up a lot of ink, but convey no data. Points can
do this more efficiently. The chart that follows has a much higher
data-to-ink ratio as it is stripped back nearly to just the data.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Dot plot of age and memory - Step 3"}

ggplot(dt, aes(Age, Memory)) +
  geom_point(size = 4) +
  theme(axis.line = element_blank())

``` 

Depending on the number of data points, one may push a bit
further. Many people in practice find they are unfamiliar with these
sort of graphs and at first it can take a bit longer to read. We are
trained and used to seeing plots with "chartjunk" and low data to
ink ratios. However, a chart like this is a far more condensed display
of data and removes distractions to really highlight the raw data or
results.

```{r, fig.width = 5, fig.height = 4, fig.cap = "Dot plot of age and memory - Step 4"}

ggplot(dt, aes(Age, Memory)) +
  geom_point(size = 4) +
  geom_text(aes(y = Memory + .5, label = Memory)) + 
  theme(axis.line = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Working memory levels across the lifespan")

``` 

Another example is the popular scatter plot. By default scatter plots
already have relatively high data to ink ratios.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 1"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point()

``` 

However, "normal" axes don't convey data, so they can be removed.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 2"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point() +
  theme(axis.line = element_blank())

``` 

If we want something *like* axes but to be more useful, we can use the
function `geom_rangeframe()` from the `ggthemes` package to put more
informative data in. Range frames add "axes" but only that go the
range of the observed data. Thus, these new axes show the minimum and
maximum of each variable.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 3"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point() +
  theme(axis.line = element_blank()) +
  geom_rangeframe() 

``` 

Finally, we can make the axis labels more informative. Instead of
presenting "pretty" numbers but that convey no data, we can pick axis
labels and breaks at meaningful points of the data.
One option is quantiles / percentiles: 0th, 25th, 50th (median), 75th
and 100th percentiles are given by default from the `quantile()`
function. Now almost every piece of ink in this figure conveys some
useful information. We can visually see the range of the `mpg` and
`hp` variables from the axes. We can see the median and interquartile
range as well[^informativeaxes].

[^informativeaxes]: These informative axes are created by asking 
`ggplot2` to create tick marks (breaks) on the x and y axis at the 
quintiles of the variables, which is the default output from the 
`quantile()` function. So now instead of the default tick marks /
breaks on the axes, the breaks and numbers are:
**minimum** (0th percentile), </br>
**25th** percentile (i.e., lower quartile), </br>
**50th** percentile (i.e., median), </br>
**75th** percentile (i.e., upper quartile), </br>
**maximum** (i.e., 100th percentile). </br>
This provides a useful descriptive statistics on each variable in the
plot, right in the axes.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 4"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point() +
  scale_x_continuous(breaks = as.numeric(quantile(mtcars$mpg))) + 
  scale_y_continuous(breaks = as.numeric(quantile(mtcars$hp))) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

``` 

In the rest of the graphs we examine, we will try to implement this
data to ink ratio principle. It does require some additional `R` code
versus simply plotting with the defaults and often at first, you may
need to spend a bit more time explaining your graph to people in text
or in the figure legend. However, ultimately, such plots convey more
data. For example, it makes it clear that cars in the top 25% of `mpg`
(i.e., 23 to 34) all fall below the median `hp` (i.e., below
123). Conversely, cares in the bottom 25% of `mpg`(i.e., 10 to 15) all
fall above the median `hp` (i.e., 123) with nearly all in the top 25%
of `hp`.

# Advanced Bivariate Graphs

As with univariate graphs, we can map additional variables to
additional aesthetics. This allows us to integrate more into standard
bivariate plots. The following graph shows a simple example of
this. 

```{r, fig.width = 6, fig.height = 5, fig.cap = "Scatter plot with shapes"}

ggplot(mtcars, aes(mpg, hp, shape = cyl)) +
  geom_point() +
  scale_x_continuous(breaks = as.numeric(quantile(mtcars$mpg))) + 
  scale_y_continuous(breaks = as.numeric(quantile(mtcars$hp))) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

However, although the shapes are distinguishable, they are
visually difficult. Cleveland did some studies around shape
perception, particularly when points may be partially overlapping and
on this basis suggested other shaped. We can manually specify the
number for which shape we want applied to which value of `cyl` using
the `scale_shape_manual()` function[^moreshapes]. We also use the
`name = ` argument so that instead of getting the legend labelled
`cyl` it is labelled `Number Cylinders`. In `R` the `\n` means
add a line break so in the legend there will be a line break between
"number" and "cylinders".

[^moreshapes]: There are many other shapes available. To see a list,
    go to: [http://sape.inf.usi.ch/quick-reference/ggplot2/shape](http://sape.inf.usi.ch/quick-reference/ggplot2/shape).

```{r, fig.width = 6, fig.height = 5, fig.cap = "Scatter plot with shapes"}

ggplot(mtcars, aes(mpg, hp, shape = cyl)) +
  geom_point(size = 3) +
  scale_shape_manual(
    name = "Number\nCylinders",
    values = c("4" = 1, "6" = 3, "8" = 83)) + 
  scale_x_continuous(breaks = as.numeric(quantile(mtcars$mpg))) + 
  scale_y_continuous(breaks = as.numeric(quantile(mtcars$hp))) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

Scatter plots are a helpful way to represent the raw data and
association between two variables. In some cases, however, it can be
harder to see the results. The following graph is a scatter plot of
stress and positive affect in the daily diary data we have used.
First we remove any missing data on those variables, then we plot.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Scatter plot with a lot of data"}

d2 <- d[!is.na(STRESS) & !is.na(PosAff)]

ggplot(d2, aes(STRESS, PosAff)) +
  geom_point(alpha = .2) +
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = as.numeric(quantile(d2$PosAff))) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

There are so many assessments and each overlap with each other so much
that it is very difficult to see any pattern or trend.
`STRESS` is skewed so both the minimum value and the 25th percentile
are 0, which is why we only see four labels on the x axis instead of
five.
We can try to better see the data by shrinking the raw points, making
them transparent and adding some noise (a.k.a., jittering) on stress.
This works relatively well for these data and helps us see the pattern
more clearly.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Scatter plot with a lot of data reducing overplotting"}

ggplot(d2, aes(jitter(STRESS), PosAff)) +
  geom_point(size = 0.6, alpha = .2) +
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = as.numeric(quantile(d2$PosAff))) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

The **breaks** for positive affect are useful but may be printed in
too much precision. We can use the `round()` function to reduce this 
precision and clean up the numbers on the y axis (e.g., if we wanted
to round to 1 decimal point).

```{r, fig.width = 6, fig.height = 5, fig.cap = "Scatter plot with a lot of data reducing overplotting plus rounding the axis breaks for positive affect"}

ggplot(d2, aes(jitter(STRESS), PosAff)) +
  geom_point(size = 0.6, alpha = .2) +
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = round(as.numeric(quantile(d2$PosAff)), digits = 1)) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

To more clearly see the trend in scatter plot data, we can add smooth
lines. `R` can add these easily using some flexible regression-based
models but that allow flexibility in the association. With repeated
measures data, the standard errors will be wrong, so we turn them off
using the argument `se = FALSE`.


```{r, fig.width = 6, fig.height = 5, fig.cap = "Scatter plot with a lot of data reducing overplotting with rounding for positive affect and a smooth trend line added"}

ggplot(d2, aes(jitter(STRESS), PosAff)) +
  geom_point(size = 0.6, alpha = .2) +
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = round(as.numeric(quantile(d2$PosAff)), digits = 2)) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe() +
  stat_smooth(se = FALSE)

```


Another option for large amounts of data where plotting the raw points
may be too much is to use summaries. For example, to examine
univariate distributions, we used histograms and density
plots. Similar options exist for bivariate data. Before we look at
some of these, we start with a 3D graph of the density of data by
stress and positive affect.

```{r, fig.width = 8, fig.height = 8, echo = FALSE, webgl=TRUE, warning=FALSE, fig.cap = "Three dimensional perspective plot of the densities of stress and positive affect"}

## This code chunk for demonstration only
## you do not need these packages or to be able to run this yourselves
library(knitr)
library(rgl)

knit_hooks$set(webgl = hook_webgl)
tmp <- MASS::kde2d(d2$STRESS, d2$PosAff)
colour <- rev(rainbow(100, start = 0/6, end = 4/6))
zcolour <- cut(tmp$z, 100)
persp3d(tmp$x, tmp$y, tmp$z, col = colour[zcolour],
        xlab = "STRESS", ylab = "PosAff", zlab = "Density")

```

Once you are familiar with this 3D representation, now we can look at
the 2D versions. One common approach is a contour plot. Each line in a
contour plot represents the same value on the third dimension, here
density. They are equally spaced, so if contours are near together, it
means that the level of the third variable changes rapidly with a
change in x or y. If they are far apart, it means the value of the
third variable changes slowly with x or y[^readingcontours].
For reference, a scatter plot of the data are also shown here. 

[^readingcontours]: 
To learn how to read contour plots, first watch this video about
reading contour lines on maps:
[https://youtu.be/CoVcRxza8nI](https://youtu.be/CoVcRxza8nI). 
The same principles apply to reading contour plots of densities,
except that instead of different lines for different altitudes, they
are different lines for different densities, which represents how many
observations fall in any particular region. </br>
Note that if data are bivariate normal, the bivariate distribution
should have a single mode (peak) that is at the mean (roughly
centered) at the mean of each variable of the bivariate
distribution. The contours would look like circles or ellipsoids
expanding out from the single peak. Multiple points of tightly spaced
contour lines may suggest more than one peak (i.e., multimodal) on one
or both dimensions. Contour lines spaced further apart indicate a
greater distance must be moved on an axis to have the same change in
density. This typically occurs when there is a more shallow slope of
the density function (e.g., a valley). Skewed distributions will tend
to have many contours lines on low or high values. Note that in a
bivariate distribution, one variable may be skewed and another normal,
or both skewed in the same or different directions.

```{r, fig.width = 6, fig.height = 5, fig.cap = "2D contour plot of densities"}

ggplot(d2, aes(STRESS, PosAff)) +
  geom_density2d(size = 1, colour = "black") +
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = round(as.numeric(quantile(d2$PosAff)), 2)) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

ggplot(d2, aes(STRESS, PosAff)) +
  geom_point() +
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = round(as.numeric(quantile(d2$PosAff)), 2)) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe() +
  ggtitle("Scatter plot for comparison")

```

Following is an example of a contour plot for bivariate normal data,
where the two variables are positive correlated with each other.

```{r, fig.width = 6, fig.height = 5, fig.cap = "2D contour plot of densities for a bivariate normal distribution and a comparison scatter plot"}

set.seed(1234)
normaldat <- data.frame(
  x = rnorm(5e4))
normaldat$y <- 2 * normaldat$x + rnorm(5e4)

ggplot(normaldat, aes(x, y)) +
  geom_density2d(size = 1, colour = "black") +
  theme(axis.line = element_blank()) +
  geom_rangeframe()

ggplot(normaldat, aes(x, y)) +
  geom_point() + 
  theme(axis.line = element_blank()) +
  geom_rangeframe() +
  ggtitle("Scatter plot for comparison")

```


We can use the `fill` argument to add colour to help. Higher numbers
indicate higher density (i.e., more observations). This is still a
contour plot, its just filled in the space in between the contour
lines with colour.

```{r, fig.width = 6, fig.height = 5, fig.cap = "2D filled contour plot of densities"}

ggplot(d2, aes(STRESS, PosAff)) +
  stat_density_2d(aes(fill = ..level..), geom = "polygon", colour = "black") + 
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = round(as.numeric(quantile(d2$PosAff)), 2)) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

One last option we will look at for large amounts of two-dimensional
data is kind of like a histogram. The data are binned, and
counted. Instead of binning on one dimension, data are binned on two
dimensions. The colour of each hexagon indicates how many people fall
into the area covered by the hexagon, just as the height of a bar in a
histogram indicates how many observations fall between the width of
the bar[^hexbin]. 

[^hexbin]: On the right hand side of the plot, the legend shows a
continuous colour gradient, where brighter blues indicate more
observations. In this instance, the brightest blues represent a count
of over 150 observations. 
The darker blues represent something under 50 observations. It will be
difficult visually to distinguish 50 observations in a bin from say 20
or 10 because all of them are a relatively dark blue. 
It is a fairly continuous colour gradient from dark to bright blue,
but visually we can only discern movement of a certain magnitude. When
you have some high frequency bins requiring that bright blues be used
for say 150+, then it gives less "space" on the colour gradient for
the low frequencies (5, 10, 20) which means all of those (visually)
appear very similar as a dark blue, and we cannot well distinguish
some bins with relatively lower frequencies apart as easily. 
Sometimes having more bins can help with this. Generally fewer bins
will exacerbate problems like this. It is also more difficult to read
these when points are very tightly clustered in a small region as that
results in some space on the graph being very high counts and others
very low counts, which is difficult to represent with a continuous
colour gradient well.

```{r, fig.width = 6, fig.height = 5, fig.cap = "2D histogram by hexagonal binning"}

ggplot(d2, aes(STRESS, PosAff)) +
  geom_hex(bins = 20) + 
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = round(as.numeric(quantile(d2$PosAff)), 2)) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

A challenge with these plots can be figuring out the ideal number of
bins to use, especially withvery skewed or "clumpy" data. Too many
people in one bin means that the colour scale must go up so high that
you do not get much granularity to compare different frequencies on
the lower end (e.g., below with `bins = 10` the colour scale is in
100s of observations). The `bins = 10` option roughly means use about
10 bins in each variable, although its used as part of an algorithm so
you don't get exactly 10 bins in the rows or columns.

```{r, fig.width = 6, fig.height = 5, fig.cap = "2D histogram by hexagonal binning"}

ggplot(d2, aes(STRESS, PosAff)) +
  geom_hex(bins = 10) + 
  scale_x_continuous(breaks = as.numeric(quantile(d2$STRESS))) + 
  scale_y_continuous(breaks = round(as.numeric(quantile(d2$PosAff)), 2)) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

```

## Try It - Bivariate

Using the daily diary dataset we have worked with all semester, `d` dataset:

1. Make a scatter plot (points) for `SUPPORT` and `COPEPrb`. Use the
   good visualization principles we have learned. Make your own
   decisions about jittering, size, and alpha transparency to make the
   scatter plot most useful to read.
2. Extend your final plot from number 1 to include an overall trend line.
3. Create a plot of the distribution/density of `SUPPORT` and
   `COPEPrb`. This can be a contour plot, hexgonal bin, or if you are
   very keen, a 3D plot.

```{r trybivariate, error=TRUE}

## scatter plot code here


## scatter plot + trend line code here


## 2d distribution plot code here


```

# Longitudinal Plots

For longitudinal data or time series data, line plots often make
sense. We have seen simple line plots before, but when there are
multiple people, using an overall line plot is not effective.

The following code subset the data to just 4 randomly sampled people
with non missing support values and creates a line plot[^randomseeds].

[^randomseeds]: The `set.seed()` function is used to set the random
seed. This controls how `R` generates random numbers. It allows
multiple computers to generate the same set of random numbers. So that
over and over, this example will randomly select the **same** random 4
IDs to plot. If you want to plot different participants, you can
change the random seed to some other number (e.g., 52349) and then you
will get a different 4 random participants' selected and plotted.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Longitudinal plots"}

set.seed(123)
d3 <- d[!is.na(SUPPORT) & UserID %in% sample(unique(d$UserID), 4)]

ggplot(d3,aes(SurveyDay, SUPPORT)) +
  geom_line()

```
The vertical jumps in the lines are because there are multiple
assessments on the same day. To address this, we need to separate the
lines by person. We can do this by using another aesthetic, such as
colour, and/or by specifying a grouping variable. We do both below.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Longitudinal plots by person"}

ggplot(d3,aes(SurveyDay, SUPPORT,
              colour = UserID, group = UserID)) +
  geom_line()

```

Just as we looked at plotting some smooth lines for scatter plots, we
can do that for each person. All we need to do is add another line
with the `stat_smooth()` function.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Longitudinal plots by person with smooths"}

ggplot(d3,aes(SurveyDay, SUPPORT,
              colour = UserID, group = UserID)) +
  geom_line() +
  stat_smooth(se=FALSE) 

```

If we think that the smooth lines are relatively good approximations,
a cleaner image may be made by just showing these, which smooths out
some of the "noise" in the raw data.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Longitudinal smooth plots by person"}

ggplot(d3,aes(SurveyDay, SUPPORT,
              colour = UserID, group = UserID)) +
  stat_smooth(se=FALSE) 

```

Finally, by default `R` uses flexible smooths. However, by specifying
`method = "lm"` we can force a linear smooth function (linear
regression), although this might be a (very) poor fit to some data,
such as ID 9.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Longitudinal linear smooth plots by person"}

ggplot(d3,aes(SurveyDay, SUPPORT,
              colour = UserID, group = UserID)) +
  geom_line() + 
  stat_smooth(method = "lm", se = FALSE) 

```


## Try It - Longitudinal

Using the daily diary dataset we have worked with all semester, `d` dataset:

1. Change the random seed and subset the dataset to just a few
   participants' IDs.
2. Plot problem focused coping over time for those few people.

```{r trylong, error=TRUE}

## subset the data


## longitudinal plot


```
