---
title: "DRAFT TBC - Graphs Part 1"
author: "Joshua F. Wiley"
date: "`r Sys.Date()`"
output: 
  tufte::tufte_html: 
    toc: true
    number_sections: true
  tufte::tufte_handout: 
    toc: true
    number_sections: true
  word_document:
    toc: false
---

```{r, echo=FALSE, include=FALSE}
## note this chunk is NOT required
library(checkpoint)
checkpoint("2019-01-25", R.version = "3.5.1",
           scanForPackages = FALSE)
```

Download the raw `R` markdown code here
[https://jwiley.github.io/MonashHonoursStatistics/Graphs_Pt1.rmd](https://jwiley.github.io/MonashHonoursStatistics/Graphs_Pt1.rmd).


```{r setup}
options(digits = 2)

## two new packages are: (1) cowplot and (2) ggthemes
## You can try installing
## by uncommenting the install.packages() code below

#install.packages("cowplot", type = "binary")
#install.packages("ggthemes", type = "binary")


## some people also report type="binary" does not work
## if that happens to you, try:
# install.packages("cowplot")
# install.packages("ggthemes")

## once installed, open the packages using the 
## library() calls below

library(data.table)
library(ggplot2)
library(cowplot)
library(ggthemes)

## read in the dataset
d <- readRDS("aces_daily_sim_processed.RDS")

## small sample dataset
dt <- data.table(
  Age = c(20, 30, 40, 50, 60),
  Memory = c(7, 5, 6, 4.5, 4))

```

# Grammar of Graphics

`ggplot2` is based on the **g**rammar of **g**raphics, a framework for
creating graphs.

The idea is that graphics or data visualization generally can be
broken down into basic low level pieces and then combined, like
language, into a final product.

Under this system, line plots and scatter plots are essentially the
same. Both have data mapped to the x and y axes.  The difference is
the plotting symbol (**ge**ometries labelled `geom`s in `R`) in
is a point or line. The data, axes, labels, titles, etc. may be
identical in both cases.

`ggplot2` also uses aesthetics, which control how geometries are
displayed. For example, the size, shape, colour, transparency level
all are **aes**thetics.

# Univariate Graphs

To begin with, we will make some simple graphs using `ggplot2`. The
first step is specifying the dataset and mapping variables to axes.
For basic, univariate plots such as a histograms, we only need to
specify the dataset and what variable is mapped to the x axis.
We can re-use this basic setup with different **geom**etries to make
different graphs.

```{r}

pb <- ggplot(
  data = mtcars,
  aes(x = mpg))

```

Using our basic mapping, we can "add" a histogram geometry to view a
histogram.

```{r, fig.width = 6, fig.height = 5, fig.cap = "A histogram in ggplot2 for mpg"}

pb + geom_histogram()

```

We also can make a density plot, which also attempts to show the
distribution, but using a smooth density function rather than binning
the data and plotting the frequencies. Like histograms, the height
indicates the relative frequency of observations at a particular
value. Density plots are designed so that they sum to one.

```{r, fig.width = 6, fig.height = 5, fig.cap = "A density plot for mpg"}

pb + geom_density()

``` 

Another type of plot are (stacked) dotplots. These are very effective
at showing raw data for small datasets. Each dot represents one
person. If two dots would overlap, they are stacked on top of each
other. While these often are difficult to view with large datasets,
for small datasets, they provide greater precision than histograms.

```{r, fig.width = 6, fig.height = 3, fig.cap = "A dot plot for mpg"}

pb + geom_dotplot()

```

## Mapping Additional Variables

We can map additional variables to aesthetics such as the colour to
include more information.  To break the dataset down further, it is
helpful to have a bit larger dataset than the 32 cars in `mtcars`. The
`iris` dataset has measurements on 150 flowers from three species of
iris. For density plots, separating by colour is easy, by adding the
`Species` variable as an additional aesthetic.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Density plot coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, colour = Species)) +
  geom_density()

```

For histograms, rather than control the colour of the lines, it is
more helpful to control the fill colour. By default, overlapping bars
are stacked on top of each other.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Histogram coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_histogram()

```

Overlapping bars also can be dodged instead of stacked.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Dodged histogram coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, fill = Species)) +
  geom_histogram(position = "dodge")

```

Multiple geometric objects can be combined. The following figure
combines dot and density plots into one. 

```{r, fig.width = 6, fig.height = 5, fig.cap = "Density and dot plot plot coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length, colour = Species)) +
  geom_dotplot() + geom_density()

```

While the colour works well for the density plots, filling would work
better for the dotplot. We can map separate aesthetics to each
geometric object.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Density and dot plot plot coloured by Species for Sepal Length"}

ggplot(iris, aes(Sepal.Length)) +
  geom_dotplot(aes(fill = Species)) +
  geom_density(aes(colour = Species))

```

## Bivariate Graphs

We can make bivariate plots by mapping variables to both the x and
y-axis. For a scatter plot, we use point geometric objects.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot of mpg and hp"}

ggplot(mtcars,
       aes(
         x = mpg,
         y = hp)) +
  geom_point()

```

We also can use lines for bivariate data. For this example, we will
use a small made up dataset. Compared to a scatter plot, we only
change point to line geometric objects.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Line plot of age and memory"}

ggplot(dt, aes(Age, Memory)) +
  geom_line()

```

With relatively discrete x axis, we can use a barplot for bivariate
data. By default, `geom_bar()` calculates the count of observations
that occur at each x value, so if we want our values to be the actual
bar height, we set `set = "identity"`.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity")

```

The grammar of graphics is designed to be like sentences, where you
can add or modify easily. For example, "There is a ball." or "There is
a big, red, striped ball." are both valid sentences. So to with
graphics, we often can chain pieces together to make it more nuanced.
In `R` we just "add" more by separating each component with `+`. 
Note that most argument names (i.e., `data = `, `mapping = `) are not
strictly required. `R` will match input to the correct argument by
position, often. The two sets of code below yield the same plot. In
the first, we explicitly label all arguments, in the second we rely on
position matching. Positional matching does not always work, for
example we still must specify `size = ` because we don't always
provide input for every argument, instead relying on defaults and only
changing the specific arguments we want changed from defaults.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory"}

ggplot(
  data = dt,
  mapping = aes(
    x = Age,
    y = Memory)) +
  geom_bar(
    mapping = aes(fill = Age),
    stat = "identity") +
  geom_line(size = 2) + 
  geom_point(size = 6)

ggplot(dt, aes(Age, Memory)) +
  geom_bar(aes(fill = Age), stat = "identity") +
  geom_line(size = 2) + 
  geom_point(size = 6)

```

# Improving Data Visualization

Before we continue examining graphs, it is helpful to think a bit more
about what makes a good graph or good data visualization.

Edward Tufte and William Cleveland are two authors who have written
extensively on data visualization and how to make good graphs. There
work is well worth reading to improving understanding on how to
efficiently convey data graphically.

- Tufte: [https://www.edwardtufte.com/tufte/](https://www.edwardtufte.com/tufte/)
- Cleveland: [http://www.stat.purdue.edu/~wsc/](http://www.stat.purdue.edu/~wsc/)

One key principle that Tufte emphasises is the data to ink ratio. This
ratio is how much data is conveyed versus ink used, and Tufte argues
to try to maximize this (i.e., more data, less ink). To see this,
consider the following graph, which is a fairly standard way stats
programs (Excel, etc.) tend to make barplots.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory - Step 1"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity") +
  theme_bw()

``` 

For starters, the borders tell us nothing. They edge the space but
convey no information. This can be cleaned up using a different
theme. Once we loaded the `cowplot` package, it automatically set its
own theme as the default, so if we leave off the `theme_bw()` we get a
cleaner graph.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory - Step 2"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity")

``` 

However, there are still some borders, which we can strip away with no
loss of data, but reducing the ink.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Bar plot of age and memory - Step 2"}

ggplot(dt, aes(Age, Memory)) +
  geom_bar(stat = "identity") +
  theme(axis.line = element_blank())

``` 

Next, think about what data are conveyed in this graph. The bars
capture two pieces of information: (1) the age and (2) the memory at
that age. The only pieces of the bars we really need are the top. The
rest of the bars take up a lot of ink, but convey no data. Points can
do this more efficiently. The chart that follows has a much higher
data-to-ink ratio as it is stripped back nearly to just the data.

```{r, fig.width = 6, fig.height = 5, fig.cap = "Dot plot of age and memory - Step 3"}

ggplot(dt, aes(Age, Memory)) +
  geom_point(size = 4) +
  theme(axis.line = element_blank())

``` 

Depending on the number of data points, one may push a bit
further. Many people in practice find they are unfamiliar with these
sort of graphs and at first it can take a bit longer to read. We are
trained and used to seeing plots with "chartjunk" and low data to
ink ratios. However, a chart like this is a far more condensed display
of data and removes distractions to really highlight the raw data or
results.

```{r, fig.width = 5, fig.height = 4, fig.cap = "Dot plot of age and memory - Step 4"}

ggplot(dt, aes(Age, Memory)) +
  geom_point(size = 4) +
  geom_text(aes(y = Memory + .5, label = Memory)) + 
  theme(axis.line = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank()) +
  ggtitle("Working memory levels across the lifespan")

``` 

Another example is the popular scatter plot. By default scatter plots
already have relatively high data to ink ratios.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 1"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point()

``` 

However, "normal" axes don't convey data, so they can be removed.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 2"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point() +
  theme(axis.line = element_blank())

``` 

If we want something *like* axes but to be more useful, we can use the
function `geom_rangeframe()` from the `ggthemes` package to put more
informative data in. Range frames add "axes" but only that go the
range of the observed data. Thus, these new axes show the minimum and
maximum of each variable.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 3"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point() +
  theme(axis.line = element_blank()) +
  geom_rangeframe() 

``` 

Finally, we can make the axis labels more informative. Instead of
presenting "pretty" numbers but that convey no data, we can pick axis
labels and breaks at meaningful points of the data.
One option is quantiles / percentiles: 0th, 25th, 50th (median), 75th
and 100th percentiles are given by default from the `quantile()`
function. Now almost every piece of ink in this figure conveys some
useful information. We can visually see the range of the `mpg` and
`hp` variables from the axes. We can see the median and interquartile
range as well.

```{r, fig.width = 5, fig.height = 5, fig.cap = "Scatter plot - Step 4"}

ggplot(mtcars, aes(mpg, hp)) +
  geom_point() +
  scale_x_continuous(breaks = as.numeric(quantile(mtcars$mpg))) + 
  scale_y_continuous(breaks = as.numeric(quantile(mtcars$hp))) + 
  theme(axis.line = element_blank()) +
  geom_rangeframe()

``` 



## TBD

```{r}

ggplot(mtcars, aes(cyl, mpg)) +
  stat_summary(fun.y = mean, geom = "bar")


ggplot(mtcars, aes(sample = mpg)) + geom_qq()

ggplot(d, aes(STRESS, PosAff)) +
  geom_point(colour = "grey50", alpha = .2)

ggplot(d, aes(STRESS, PosAff)) +
  geom_jitter(colour = "grey50", alpha = .2) +
  geom_density2d(size = 1, colour = "black") 

```
