---
title: "My First R Markdown File"
author: "Joshua F. Wiley"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: 
    toc: true
    number_sections: true
---

Download the raw `R` code here:
[https://jwiley.github.io/MonashHonoursStatistics/MyFirstRMarkdown.rmd](https://jwiley.github.io/MonashHonoursStatistics/MyFirstRMarkdown.rmd).

# `R` Markdown Files

`R` markdown files combine `R` code for data analysis with markdown, a
markup language that adds some additional text to plain text to manage
formating. For example, asterisks can be used around words to
*italicize* the words, or double asterisks to **make words bold**.
Hashtags create headings, with one, two, three, etc. hashtags
creating:

# Level 1 Heading

## Level 2 Heading

### Level 3 Heading

The benefit of `R` and markdown together is that you can write `R`
code and in between your code, you can write regular text in markdown
to explain what the code does or interpret the output.
To let the computer know some of what we type is `R` code, we
delimitters that indicate the start and end of `R` code: triple
backticks and `r`. Here is our first `R` code chunk, where we ask `R`
to calculate 6 * 5.5. In the markdown part, nothing happens, but when
written as `R` code it can be processed and `R` will work as a
calculator.

```{r}

6 * 5.5

```

The whole set of `R` code above is called a "chunk" and any document
can have none, one or more `R` code chunks. Inside the curly braces
that start each chunk, we can write additional code to give the chunk
a name and control other aspects of how it is run. For example, let's
call the next chunk "TestName" and there let's find out the homework
mark if someone did 14 of the chapters and 4 of the courses from the
homework software exercises.

```{r TestName}

14 * (1/46) + 4 * (4/46)


```

# Using `R` Packages

We often want to do more data analysis or use tools that are not part
of "default" `R`. There are many packages (like apps for a phone) that
augment `R`. In fact there are over 10,000 such packages. The 
[setuppage](http://joshuawiley.com/MonashHonoursStatistics/IntroR.html) 
linked in Week 1 for Moodle had you install several of these.
In our first `R` markdown file, let's just use two important ones:

- `data.table` is a package with extensive data management tools
- `ggplot2` is the second version of the **g**rammar of **g**raphics
  plot package. It makes amazing graphs in `R` and we'll rely on it
  for data visualization.
  
To open a package (app), instead of tapping like a phone, we use the
`library()` function. In `R` functions always have a name and then an
open and close paretheses. We add arguments or options inside the
parentheses to tell the funciton what we want it to do specifically.
The `library()` function is used to access our library of packages and
open a specific one, and the argument we need to give it is the name.
Here is an example.


```{r}

## open the data.table app for data management
library(data.table)

## open the ggplot2 app for graphing
library(ggplot2)

```

In the `R` chunk above, you might have noticed that hash tags were
used, but they did not make level 2 headings. Inside an `R` chunk,
what we write behaves *differently* to outside and `R` chunk. In `R`
hashtags indicate that everything that follows on that line is a
comment. That is, its not actually code, its just a short note or
comment to ourselves or our reader about what this code is supposed to
do. Now, let's do a few simple things with a dataset.
First, we'll convert the built in `mtcars` dataset with different
features of 32 cars into a `data.table` and we'll call it "d" 
(formally, we assign the results of the function call to an object,
"d"). That means when we type "d" in the future, in `R` that will mean
the data.table of the mtcars data. Then we'll look at a few rows of
the data, and view only the rows that correspond to cars that get over
30 miles per gallon (this is an old dataset).


```{r}

## convert mtcars dataset to a data.table
d <- as.data.table(mtcars)

## view the first few rows of the data
head(d)

## view only cars that get over 30 mpg
d[mpg > 30]

```

Finally, let's take a look at a couple basic graphs using `ggplot2`.
We'll cover how to create graphs and visualize data in more depth
later, for now, just focus on how to run `R` code and where the code,
output and explanation live in an `R` markdown file.

```{r}

## bar plot
ggplot(d, aes(cyl)) + 
  geom_bar()

```

Sometimes, we might want to adjust the size of the graph. We can do
this through *chunk options* which are addded in the header of the chunk.

```{r, fig.width = 6, fig.height = 4}

## bar plot
ggplot(d, aes(cyl)) + 
  geom_bar()

```

The very last step is to turn all of this code into a document that
you can reference later or could send to someone else. This is one of
the great parts of using `R` markdown files. You may do some analysis
now and need to refer to it later, or you may do some analysis and
want to send it to a coworker who knows nothing about `R` but you
still want them to see all yours notes and all the output. We'll do
this by `render`ing the `R` markdown file. You can do this through
code, but we'll just use `RStudio` to help us do it.
