---
title: "PSY4210 Lab Report - 2023 - Part 1"
author: "Your Monash ID Number -- Your Full Name"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: 
    toc: true
---

# 1. Generalized Linear Models 1


```{r}

library(haven)
library(data.table)
library(JWileymisc)
library(ggplot2)
library(ggpubr)
library(visreg)


## read in data
db <- as.data.table(read_sav("[2021] PSY4210 BL.sav")) # baseline data

```

## Run a Linear Regression

Fit **one** moderated, multiple linear regression. Pick from:

- Outcome: Anxiety at baseline
- Predictors: `sex` and any other variable of your choosing
- Make sure that you include the interaction between your predictor and `sex`.

Note: you will need to score Anxiety from the STAI. For a
refresher, see the content for Data Visualization 1.

```{r glm1reg}

## put your regression model code using lm() here



```

## Linear Regression Diagnostics

Conduct model diagnostics on your regression model. Where applicable,
apply appropriate transformations and/or exclude outliers.

```{r glm1diagnostics}

## put your regression diagnostics code including plots using modelDiagnostics() here



```


## Linear Regression Summary

Create a summary of your **final** model, after any relevant
transformations or extreme values have been addressed, where
applicable.

- use `modelTest()` and `APAStyler()` to get a nice result.

Create a graph to help visualize your results using `visreg()`.
Note, you may want to adjust the `fig.width` and `fig.height` options
to this `R` chunk to make the graph have the ratios you like.

```{r glm1plot, fig.width = 6, fig.height = 4}

## put your code to summarize the final model using modelTest() here


## put your code to visualize your regression model using visreg() here


```

In the last subheading (Linear Regression Write Up) for GLM1, briefly
summarize your analysis steps, assumption checks, any changes
performed, and interpret the interaction and regression coefficients,
referencing your figure visualizing the result where appropriate. 
The total length should be about one paragraph and about 150 words.

## Linear Regression Write Up




# 2. Generalized Linear Models 2



